Modeling the Latent Space of Symbolic String Quartet Music

### 2021-06-08 papers/final/alex/paper.tex
write discussion section of paper

Evaluation of generative models is challenging because there is no
direct equivalent of an accuracy or error metric like what is used in
supervised learning. For autoencoder models we can measure how
accurately the model can reconstruct its own inputs, but this does not
tell us the quality of the interpolated examples. Generative models
are typically evaluated using a combination of qualitative metrics
whereby human judges rate the quality of the generated examples and
guess whether they were generated by a machine (essentially a Turing
test), and quantitative metrics that assess the differences in the
parametric distributions of generated and real examples. Yang and
Lerch (2020) propose a set of metrics informed by music theory, for
probabilistically evaluating how similar the generations are to known
sample distributions of real music \cite{yang_evaluation_2020}. These
metrics include counts, ranges, histograms and transition matrices of
pitches and note lengths, then the Kullback-Leibler divergence and
overlapping area of the probability density or mass functions are
used to compare against known reference distributions per musical
genre \cite{yang_evaluation_2020}. Due to the cost and time
requirements associated with designing a human subjects experiment for
qualitative evaluation, we utilize this quantitative approach to the
assessment of generated samples.

@@

@@

@@

@@

>>>>

Evaluation of generative models is challenging because there is no
direct equivalent of an accuracy or error metric like what is used in
supervised learning. For autoencoder models we can measure how
accurately the model can reconstruct its own inputs, but this does not
tell us the quality of the interpolated examples. Generative models
are typically evaluated using a combination of qualitative metrics
whereby human judges rate the quality of the generated examples and
guess whether they were generated by a machine (essentially a Turing
test), and quantitative metrics that assess the differences in the
parametric distributions of generated and real examples. Yang and
Lerch (2020) propose a set of metrics informed by music theory, for
probabilistically evaluating how similar the generations are to known
sample distributions of real music \cite{yang_evaluation_2020}. These
metrics include counts, ranges, histograms and transition matrices of
pitches and note lengths, then compare the probability density or mass
functions of generated samples against that of known genuine source
samples \cite{yang_evaluation_2020}. Due to the cost and time
requirements associated with designing a human subjects experiment for
qualitative evaluation, we utilize this quantitative approach to the
assessment of the pitch distributions of generated samples.

@@

Figures \ref{source_ranges} and \ref{generated_ranges} depict the
distribution of pitch range (highest pitch - lowest pitch) among
samples. The distributions appear similar; source distribution has a
mean of 30.5 while the generated distribution has a mean of 26.3, (the
medians are similar to the means for both) so the generated
distributions have a slightly narrower pitch range.

@@

Figures \ref{source_interval} and \ref{generated_interval} describe
the distributions average pitch intervals between consecutively played
notes; these distributions are very similar, both are centered with
approximate mean of 0 and standard deviation of 1.5.

@@

Finally, figures \ref{pctm_source} and \ref{pctm_generated} depict
heatmaps of transition probability between the twelve pitch classes in
the chromatic scale. The source samples exhibit higher probabilities
of transition between different notes, while in the generated sample
there is more repetition of the same note, again confirming the
relative lack of pitch diversity in the generated samples. However,
the dark areas of very low probability in the source distribution are
similarly dark in the generated distribution; the model is unlikely to
generate pitch transitions that are rare in the source distribution.

@@

Our research demonstrates that a relatively simple LSTM variational
autoencoder network can learn a continuous latent space for generating
realistic linear interpolations between symbolic, classical string
quartet pieces, which are a notably complex genre of music.
Furthermore, our model can achieve this while working
measure-by-measure, on sequences just 16 time steps in length, without
the need for a highly complex neural architecture capable of
memorizing very long sequences. Listening to the generated samples and
profiling their pitch distributions reveals that some of the pitch
diversity of the original music is lost, but the overall melodic and
rhythmic structure still generally resemble the originals. It was our
aim to produce interpolations that retained enough aesthetic quality
to be pleasing to listen to and useful in inspiring the creative
process, and we believe our work meets that goal.

Future work in this domain could involve disentanglement of the latent
dimensions to enable music style transfer and semantic editing via
interpolation in specific directions within the latent space. The VAE
model could be adapted to other musical genres aside from string
quartets, to interpolate within or across genres. The model could also
be modified into a conditional VAE to allow conditioning on the genre,
composer/artist name, or some other defining attribute label, to
influence the generated results.
